---
title: 'Storms: Human and Property Damage Analysis'
output: pdf_document
---

```{r load_libraries, echo=FALSE, message=FALSE} 
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```

```{r load_data, echo=FALSE}

# storm.data <- tbl_df(
#   read.csv(file = "data//StormData.csv", header = TRUE, #nrows = 1000,
#            colClasses = c(
#              "NULL", 
#              "character", #date
#              rep("NULL", times = 4),
#              "factor", #state
#              "factor", #type
#              rep("NULL", times = 14),
#              "numeric", #fatal
#              "numeric", #inju
#              "numeric", #property
#              rep("NULL", times = 12)
#           )
#   )
# )
# 
# 
# saveRDS(object = storm.data, ascii = FALSE, compress = "bzip2", file = "data//stormdata.rds")

storm.data <- readRDS(file = "data//stormdata.rds")

```

```{r manip_data, echo=FALSE}
storm.data2 <- storm.data %>% 
  transmute(
    date = lubridate::mdy(
      x = sub(x = BGN_DATE, pattern = "([0-9/]+).*", replacement = "\\1")),
    decade = as.factor(10 * (year(date) %/% 10)),
    EVTYPE, FATALITIES, INJURIES, PROPDMG
)
```

Afterwards the categories became more open form. In 1993 there were 160 categories,
1994: 267, 1995: 387, 1996: 228. However, the distinctions were often small, 
for example there were three entries
for "WIND CHILL/HIGH WIND", "HIGH WIND/WIND CHILL/BLIZZARD", "HIGH WIND/WIND CHILL"
as well as separate events sometimes labeled by their date, or occasionally by 
the name of a specific storm.

```` {r shrink_data, echo=FALSE}
impact <- storm.data2 %>%
  filter(decade %in% c(2000, 2010)) %>%
  group_by(EVTYPE) %>%
  summarise(
    death.mean = mean(FATALITIES), death.tot = sum(FATALITIES), 
    injuries.mean = mean(INJURIES), injuries.tot = sum(INJURIES),
    property.mean = mean(PROPDMG), property.tot = sum(PROPDMG), 
    denom = n()
    )
```

For the sake of analysis, I will choose a few categories:

* Tornado
* Hurricane/Typhoon
* Heat
* Flooding

```{r}


```

